---
title: "Draft Mosaic Benchmark Database"
author: "Adetola Abdulkadir"
date: '`r Sys.Date()`'
output: 
    bookdown::html_document2:
        toc: true
        toc_float: true
        df_print: paged
        code_folding: hide
---

# Objective
Creation of a database for the draft benchmark variants. This database includes performance of 4
orthogonal methods: Duplex Sequencing, Element, BGI, and Pacbio HiFi. The database will also
include annotations that categorize variants, highlight features (such as SomaticEVS), and help
us identify which variants should be manually curated.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(knitr)
library(stringr)
library(dplyr)
library(gt)
library(Hmisc)
```

# Download and cleaning of the draft benchmark

## Download
```{r draft benchmark upload}
here::i_am('scratch/draft_mosaic_benchmark.Rmd')
draft_vcf <- read_tsv(here("data", "cand_put_mosaic_v0.0.3.FINAL.vcf"), comment = "##") %>%
    rename(CHROM = `#CHROM`)
# SEE README FOR DESCRIPTION ON BENCHMARK GENERATION
```

## Annotate variants
```{r}
var_type <- draft_vcf %>%
    mutate(var_cat = case_when(
        FILTER == "PASS" ~ "candidate",
        TRUE             ~ "putative"),
        .after = FILTER
    ) %>%
    mutate(
        REF_len = nchar(REF),
        ALT_len = nchar(ALT)
    ) %>%
    mutate(var_type = case_when(
        REF_len == 1 & ALT_len == 1 ~ "snv",
        REF_len == 1 & ALT_len > 1  ~ "indel",
        REF_len > 1  & ALT_len == 1 ~ "indel",
        REF_len > 1  & ALT_len > 1  ~ "MNP",
        TRUE                         ~ "FIX_ME"
    ), .after = var_cat) %>%
    select(-ID, -QUAL)

var_type %>%
    group_by(var_cat, var_type) %>%
    dplyr::summarize(n = n()) %>%
    ungroup()

snv_info_cols <- c("SOMATIC","QSS","TQSS","NT","QSS_NT","TQSS_NT",
                   "SGT","DP","MQ","MQ0","ReadPosRankSum","SNVSB","SomaticEVS",
                   "var_id","dup_var_hit","easy_to_map","homopolymer_region",
                   "germline_insertion","germline_deletion")

snv_fmt_value  <- "DP:FDP:SDP:SUBDP:AU:CU:GU:TU:GT"
snv_fmt_cols   <- c("DPtier1","FDP","SDP","SUBDP","AU","CU","GU","TU","GT")
snv_numeric_cols <- c("QSS","TQSS","QSS_NT","TQSS_NT","MQ","MQ0",
                      "ReadPosRankSum","SNVSB","SomaticEVS","DP","FDP","SUBDP")

draft_snvs <- draft_vcf %>%
    filter(FORMAT == snv_fmt_value) %>%
    separate(HG002, into = snv_fmt_cols,  sep = ":") %>%
    separate(INFO,  into = snv_info_cols, sep = ";") %>%
    select(-FORMAT, -ID, -QUAL) %>%
    mutate(across(.cols = all_of(snv_info_cols),    str_remove, ".*=")) %>%
    mutate(across(.cols = all_of(snv_numeric_cols), as.numeric)) %>%
    left_join(var_type)
```

### Sanity check variant counts
```{r}
draft_snvs %>%
    filter(var_cat == 'putative') %>%
    mutate(SomaticEVS_Category = case_when(
        SomaticEVS == 0 ~ '0_EVS',
        SomaticEVS <= 1 ~ '0to1_EVS',
        SomaticEVS <= 3 ~ '1to3_EVS',
        SomaticEVS < 7  ~ '3to7_EVS',
        TRUE            ~ '7plus_EVS'
    )) %>%
    group_by(SomaticEVS_Category) %>%
    dplyr::summarize(n = n()) %>%
    gt() %>%
    tab_header(title = paste("Somatic EVS of Putative SNVs (Total n =",
                             nrow(filter(draft_snvs, var_cat == 'putative')), ")"))

draft_snvs %>%
    group_by(var_cat) %>%
    dplyr::summarize(n = n()) %>%
    gt() %>%
    tab_header(title = paste('Number of SNVs within Draft Benchmark (Total n =',
                             nrow(draft_snvs), ')'))
```

## Calculating strelka AF for SNVs
```{r}
draft_snvs_af_df <- draft_snvs %>%
    mutate(
        refCounts = case_when(
            REF == "A" ~ AU, REF == "C" ~ CU,
            REF == "G" ~ GU, REF == "T" ~ TU,
            TRUE        ~ "-1,-1"
        ),
        altCounts = case_when(
            ALT == "A" ~ AU, ALT == "C" ~ CU,
            ALT == "G" ~ GU, ALT == "T" ~ TU,
            TRUE        ~ "-1,-1"
        )
    ) %>%
    separate(refCounts, into = c("tier1RefCounts", "tier2RefCounts"), convert = TRUE) %>%
    separate(altCounts, into = c("tier1AltCounts", "tier2AltCounts"), convert = TRUE) %>%
    mutate(
        AltCounts = if_else(TQSS_NT == 1, tier1AltCounts, tier2AltCounts),
        RefCounts = if_else(TQSS_NT == 1, tier1RefCounts, tier2RefCounts),
        AF        = AltCounts / (AltCounts + RefCounts)
    ) %>%
    select(CHROM, POS, var_type, var_cat, SomaticEVS, REF, ALT, AF,
           var_id, dup_var_hit, easy_to_map, homopolymer_region,
           germline_insertion, germline_deletion)
```

### Sanity Check variant counts
```{r}
draft_snvs_af_df %>%
    mutate(VAF_Category = case_when(
        AF == 0    ~ '0% VAF',
        AF <= 0.01 ~ '0% - 1% VAF',
        AF <= 0.04 ~ '1% - 4% VAF',
        AF < 0.05  ~ '4% - 5% VAF',
        AF < 0.1   ~ '5% - 10% VAF',
        AF < 0.3   ~ '10% - 30% VAF',
        TRUE       ~ '>30% VAF'
    )) %>%
    group_by(VAF_Category) %>%
    dplyr::summarize(n = n()) %>%
    gt() %>%
    tab_header(title = 'Variant Allele Frequency (VAF) of SNVs')
```

## Calculating strelka AF for Indels
```{r}
indel_fmt_value    <- "DP:DP2:TAR:TIR:TOR:DP50:FDP50:SUBDP50:BCN50:GT"
indel_fmt_cols     <- c("DP","DP2","TAR","TIR","TOR","DP50","FDP50","SUBDP50","BCN50","GT")
indel_info_cols    <- c("SOMATIC","QSI","TQSI","NT","QSI_NT","TQSI_NT",
                        "SGT","MQ","MQ0","RU","RC","IC","IHP","SomaticEVS",
                        "var_id","dup_var_hit","easy_to_map","homopolymer_region",
                        "germline_insertion","germline_deletion")
indel_numeric_cols <- c("QSI","TQSI","QSI_NT","TQSI_NT","MQ","MQ0","RC","IC",
                        "IHP","SomaticEVS","DP","DP2","DP50","FDP50","SUBDP50","BCN50")

draft_indels <- draft_vcf %>%
    filter(FORMAT == indel_fmt_value) %>%
    separate(HG002, into = indel_fmt_cols,  sep = ":") %>%
    separate(INFO,  into = indel_info_cols, sep = ";") %>%
    select(-FORMAT) %>%
    mutate(across(.cols = all_of(indel_info_cols),    str_remove, ".*=")) %>%
    mutate(across(.cols = all_of(indel_numeric_cols), as.numeric)) %>%
    left_join(var_type)
```

### Sanity check variant counts
```{r}
draft_indels %>%
    filter(var_cat == 'putative') %>%
    mutate(SomaticEVS_Category = case_when(
        SomaticEVS == 0 ~ '0_EVS',
        SomaticEVS <= 1 ~ '0to1_EVS',
        SomaticEVS <= 3 ~ '1to3_EVS',
        SomaticEVS < 7  ~ '3to7_EVS',
        TRUE            ~ '7plus_EVS'
    )) %>%
    group_by(SomaticEVS_Category) %>%
    dplyr::summarize(n = n()) %>%
    gt() %>%
    tab_header(title = paste('Somatic EVS of Putative Indels (Total n =',
                             nrow(filter(draft_indels, var_cat == 'putative')), ')'))

draft_indels %>%
    group_by(var_cat) %>%
    dplyr::summarize(n = n()) %>%
    gt() %>%
    tab_header(title = paste('Number of Indels within Draft Benchmark (Total n =',
                             nrow(draft_indels), ')'))
```

## Continue indel AF calculation
```{r}
draft_indels_af_df <- draft_indels %>%
    separate(TAR, into = c("tier1RefCounts", "tier2RefCounts"), convert = TRUE) %>%
    separate(TIR, into = c("tier1AltCounts", "tier2AltCounts"), convert = TRUE) %>%
    mutate(
        AltCounts = if_else(TQSI_NT == 1, tier1AltCounts, tier2AltCounts),
        RefCounts = if_else(TQSI_NT == 1, tier1RefCounts, tier2RefCounts),
        AF        = AltCounts / (AltCounts + RefCounts),
        total     = tier1AltCounts + tier2AltCounts + tier1RefCounts + tier2RefCounts
    ) %>%
    select(CHROM, POS, var_type, var_cat, SomaticEVS, REF, ALT, AF,
           var_id, dup_var_hit, easy_to_map, homopolymer_region,
           germline_insertion, germline_deletion) %>%
    mutate(indel_change = case_when(
        str_length(REF) > str_length(ALT) ~ str_c('-', str_remove(REF, ALT)),
        str_length(REF) < str_length(ALT) ~ str_c('+', str_remove(ALT, REF))
    ), .after = ALT)
```

### Sanity check
```{r}
draft_indels_af_df %>%
    mutate(VAF_Category = case_when(
        AF == 0    ~ '0% VAF',
        AF <= 0.01 ~ '0% - 1% VAF',
        AF <= 0.04 ~ '1% - 4% VAF',
        AF < 0.05  ~ '4% - 5% VAF',
        AF < 0.1   ~ '5% - 10% VAF',
        AF < 0.3   ~ '10% - 30% VAF',
        TRUE       ~ '>30% VAF'
    )) %>%
    group_by(VAF_Category) %>%
    dplyr::summarize(n = n()) %>%
    gt() %>%
    tab_header(title = 'Variant Allele Frequency (VAF) of Indels')
```

## Combine indels and snvs
```{r}
draft_variants <- draft_indels_af_df %>%
    select(-indel_change) %>%
    {rbind(draft_snvs_af_df, .)} %>%
    mutate(dup_var_hit = case_when(
        var_id == 'hit' ~ 'hit',
        TRUE            ~ dup_var_hit
    )) %>%
    mutate(var_id = case_when(
        var_id == 'hit' ~ NA,
        TRUE            ~ var_id
    )) %>%
    mutate(indel_change = case_when(
        str_length(REF) > str_length(ALT) ~ str_c('-', str_remove(REF, ALT)),
        str_length(REF) < str_length(ALT) ~ str_c('+', str_remove(ALT, REF))
    ), .after = ALT)
```

# Including HG002 DupSeq data into the database

## Downloading the HG002 Duplex Sequencing data
```{r}
mut_metadata <- read_tsv(here("data/dupseq-file-metadata.tsv"), comment = "#") %>%
    filter(File_type == "mut" & grepl("GIAB-HG002", sample) == TRUE)
mut_list <- as.list(mut_metadata$url) %>% set_names(mut_metadata$sample)
```

## Detecting benchmark variants within HG002
```{r}
tidy_and_annotate_mut <- function(filepath, target_metadata){
    mut_tbl <- read_tsv(filepath)

    mut_tbl <- mut_tbl %>%
        select(contig, start, end, variation_type,
               ref, alt, alt_depth, depth, no_calls, qual)

    mut_tbl <- target_metadata %>%
        rename(contig = CHROM) %>%
        left_join(mut_tbl, by = join_by(contig, POS == end, REF == ref, ALT == alt))

    return(mut_tbl)
}

HG002_mut_tbl <- mut_list %>%
    map_dfr(tidy_and_annotate_mut, draft_variants, .id = "sample_id") %>%
    mutate(MAF = alt_depth / (depth - no_calls)) %>%
    mutate(method = "Duplex")

HG002_mut_tbl_depth <- HG002_mut_tbl %>%
    select(-variation_type, -alt_depth, -no_calls, -MAF, -qual, -start) %>%
    pivot_wider(names_from = sample_id, values_from = depth)

HG002_mut_tbl_VAF <- HG002_mut_tbl %>%
    select(-variation_type, -depth, -alt_depth, -no_calls, -qual, -start) %>%
    pivot_wider(names_from = sample_id, values_from = MAF)

HG002_mut_tbl_combined <- HG002_mut_tbl %>%
    select(-sample_id) %>%
    rename('CHROM' = 'contig') %>%
    group_by(CHROM, POS, var_type, var_cat, var_id, dup_var_hit, easy_to_map,
             homopolymer_region, germline_insertion, germline_deletion,
             SomaticEVS, REF, ALT, AF, method) %>%
    summarise(
        Duplex_alt_depth = sum(alt_depth, na.rm = TRUE),
        Duplex_depth     = sum(depth,     na.rm = TRUE),
        Duplex_no_calls  = sum(no_calls,  na.rm = TRUE),
        Duplex_qual      = median(qual,   na.rm = TRUE),
        .groups = 'drop'
    ) %>%
    mutate(Duplex_mut_freq = Duplex_alt_depth / (Duplex_depth - Duplex_no_calls)) %>%
    mutate(Duplex_Lower_CI = binconf(Duplex_alt_depth, Duplex_depth - Duplex_no_calls, alpha = 0.10)[2]) %>%
    mutate(Duplex_Upper_CI = binconf(Duplex_alt_depth, Duplex_depth - Duplex_no_calls, alpha = 0.10)[3])
```

# Downloading Orthogonal Data

## Helper functions for tidy,annotating, and matching orthogonal variant data to benchmark
```{r helpers}

bam_readcount_col_types <- cols(
    map_qual_A = 'd', map_qual_G = 'd', map_qual_T = 'd', map_qual_C = 'd',
    base_qual_A = 'd', base_qual_G = 'd', base_qual_T = 'd', base_qual_C = 'd'
)

annotation_cols <- c("var_id","dup_var_hit","easy_to_map","homopolymer_region",
                     "germline_insertion","germline_deletion")

# Key columns shared across every platform output tibble.
variant_key_cols <- c("CHROM","POS","var_type","var_cat","SomaticEVS","REF","ALT","AF",
                      annotation_cols)

tidy_and_annotate_orth_mut <- function(mut_tbl, target_metadata, mut_type) {

    if (mut_type == 'snv') {
        mut_tbl <- left_join(target_metadata, mut_tbl,
                             by = c('CHROM' = 'chrom', 'POS' = 'position')) %>%
            select(-ref_base, -alt_base) %>%
            mutate(method_alt_depth = case_when(
                ALT == 'A' ~ field_A, ALT == 'C' ~ field_C,
                ALT == 'T' ~ field_T, ALT == 'G' ~ field_G
            )) %>%
            mutate(method_map_qual = case_when(
                ALT == 'A' ~ map_qual_A, ALT == 'C' ~ map_qual_C,
                ALT == 'T' ~ map_qual_T, ALT == 'G' ~ map_qual_G
            )) %>%
            mutate(method_base_qual = case_when(
                ALT == 'A' ~ base_qual_A, ALT == 'C' ~ base_qual_C,
                ALT == 'T' ~ base_qual_T, ALT == 'G' ~ base_qual_G
            )) %>%
            rename('method_depth' = 'depth') %>%
            mutate(method_mut_freq = method_alt_depth / method_depth) %>%
            select(-alt_depth)
        return(mut_tbl)
    }

    if (mut_type == 'indel') {
        mut_tbl <- left_join(target_metadata, mut_tbl,
                             by = c('CHROM' = 'chrom', 'POS' = 'position'))

        gathered_mut_tbl <- mut_tbl %>%
            # gather() replaced with pivot_longer() (gather was deprecated in tidyr 1.0)
            pivot_longer(
                cols      = starts_with("field_indel"),
                names_to  = "indel_column",
                values_to = "indel_value"
            ) %>%
            separate(col  = 'indel_value',
                     into = c('indel_type','indel_val','method_map_qual','method_base_qual'),
                     sep  = ":") %>%
            filter(indel_change == indel_type) %>%
            select(-ref_base, -alt_base, -alt_depth, -mut_freq,
                   -field_A, -field_G, -field_C, -field_T, -field_N,
                   -indel_column, -indel_type,
                   -map_qual_A, -map_qual_C, -map_qual_G, -map_qual_T,
                   -base_qual_A, -base_qual_C, -base_qual_G, -base_qual_T) %>%
            # Explicit join keys replace the original implicit right_join(target_metadata)
            right_join(target_metadata, by = c("CHROM","POS","REF","ALT")) %>%
            mutate(method_alt_depth  = as.numeric(indel_val)) %>%
            rename('method_depth' = 'depth') %>%
            mutate(method_mut_freq  = method_alt_depth / method_depth) %>%
            select(-indel_val) %>%
            mutate(method_map_qual  = as.numeric(method_map_qual),
                   method_base_qual = as.numeric(method_base_qual))

        return(gathered_mut_tbl)
    }
}

read_and_annotate_platform <- function(filepath, target_metadata, mut_type,
                                        platform_name, extra_col_types = cols()) {
    col_spec <- modifyList(bam_readcount_col_types, extra_col_types)
    read_tsv(filepath, col_types = col_spec) %>%
        tidy_and_annotate_orth_mut(target_metadata, mut_type) %>%
        rename_with(
            ~ str_replace(., "^method_", paste0(platform_name, "_")),
            starts_with("method_")
        ) %>%
        mutate(method = platform_name)
}

combine_platform_variants <- function(snv_target, indel_target, platform_name) {
    p         <- platform_name
    depth_col <- paste0(p, "_depth")
    alt_col   <- paste0(p, "_alt_depth")
    map_col   <- paste0(p, "_map_qual")
    base_col  <- paste0(p, "_base_qual")
    freq_col  <- paste0(p, "_mut_freq")
    lower_col <- paste0(p, "_Lower_CI")
    upper_col <- paste0(p, "_Upper_CI")

    platform_cols <- c(depth_col, alt_col, map_col, base_col, freq_col)

    combined <- bind_rows(
        select(snv_target,   all_of(c(variant_key_cols, platform_cols, "method"))),
        select(indel_target, all_of(c(variant_key_cols, platform_cols, "method")))
    ) %>%
        mutate(
            .x_tmp = coalesce(.data[[alt_col]],   0),
            .n_tmp = coalesce(.data[[depth_col]], 0)
        )

    ci <- binconf(combined$.x_tmp, combined$.n_tmp, alpha = 0.10)

    combined %>%
        bind_cols(as_tibble(ci)) %>%
        rename(!!lower_col := Lower, !!upper_col := Upper) %>%
        select(-.x_tmp, -.n_tmp, -PointEst) %>%
        mutate(index = row_number())
}

summarise_reads <- function(..., platform_name) {
    p         <- platform_name
    alt_col   <- paste0(p, "_alt_depth")
    depth_col <- paste0(p, "_depth")
    map_col   <- paste0(p, "_map_qual")
    base_col  <- paste0(p, "_base_qual")
    freq_col  <- paste0(p, "_mut_freq")

    bind_rows(...) %>%
        mutate(method = platform_name) %>%
        group_by(across(all_of(c(variant_key_cols, "method")))) %>%
        summarise(
            !!alt_col   := sum(.data[[alt_col]],       na.rm = TRUE),
            !!depth_col := sum(.data[[depth_col]],     na.rm = TRUE),
            !!map_col   := median(.data[[map_col]],    na.rm = TRUE),
            !!base_col  := median(.data[[base_col]],   na.rm = TRUE),
            .groups     = 'drop'
        ) %>%
        mutate(!!freq_col := .data[[alt_col]] / .data[[depth_col]])
}

rename_platform_prefix <- function(tbl, from, to) {
    tbl %>% rename_with(
        ~ str_replace(., paste0("^", from, "_"), paste0(to, "_")),
        starts_with(paste0(from, "_"))
    )
}
```

## Downloading BGI Data
```{r}
BGI_snv_target <- read_and_annotate_platform(
    here("scratch","bam_readcount","BGI_RM_snv.tsv"),
    draft_snvs_af_df, 'snv', 'BGI'
)
BGI_indel_target <- read_and_annotate_platform(
    here("scratch","bam_readcount","BGI_RM_indel.tsv"),
    draft_indels_af_df, 'indel', 'BGI'
)
BGI_variants_target <- combine_platform_variants(BGI_snv_target, BGI_indel_target, 'BGI')
```

## Downloading Element Data
```{r}
Element_snv_standard_target <- read_and_annotate_platform(
    here("scratch","bam_readcount","Element_RM_snv_standard.tsv"),
    draft_snvs_af_df, 'snv', 'Element'
)
Element_snv_long_target <- read_and_annotate_platform(
    here("scratch","bam_readcount","Element_RM_snv_long.tsv"),
    draft_snvs_af_df, 'snv', 'Element'
)
Element_indel_standard_target <- read_and_annotate_platform(
    here("scratch","bam_readcount","Element_RM_indel_standard.tsv"),
    draft_indels_af_df, 'indel', 'Element'
)
Element_indel_long_target <- read_and_annotate_platform(
    here("scratch","bam_readcount","Element_RM_indel_long.tsv"),
    draft_indels_af_df, 'indel', 'Element'
)

# Combined across standard and long reads
Element_snv_combined   <- summarise_reads(Element_snv_standard_target, Element_snv_long_target,
                                          platform_name = 'Element')
Element_indel_combined <- summarise_reads(Element_indel_standard_target, Element_indel_long_target,
                                          platform_name = 'Element')
Element_variants_target_combined <- combine_platform_variants(
    Element_snv_combined, Element_indel_combined, 'Element'
)

# Sub-instrument views (standard and long read sets separately)
Element_variants_target_combined_long <- combine_platform_variants(
    rename_platform_prefix(Element_snv_long_target,   "Element", "Element_long"),
    rename_platform_prefix(Element_indel_long_target, "Element", "Element_long"),
    'Element_long'
)
Element_variants_target_combined_standard <- combine_platform_variants(
    rename_platform_prefix(Element_snv_standard_target,   "Element", "Element_standard"),
    rename_platform_prefix(Element_indel_standard_target, "Element", "Element_standard"),
    'Element_standard'
)
```

## Downloading Pacbio Data
```{r}
# Extra col types specific to Revio indel files
revio_indel_extra <- cols(position = 'd', alt_depth = 'd', depth = 'd',
                          mut_freq = 'd', field_A = 'd')

Pacbio_revio_snv_target <- read_and_annotate_platform(
    here("scratch","bam_readcount","Pacbio_RM_revio_snv.tsv"),
    draft_snvs_af_df, 'snv', 'Pacbio'
)
Pacbio_revio_indel_target <- read_and_annotate_platform(
    here("scratch","bam_readcount","Pacbio_RM_revio_indel.tsv"),
    draft_indels_af_df, 'indel', 'Pacbio',
    extra_col_types = revio_indel_extra
)
Pacbio_sequel_snv_A_target <- read_and_annotate_platform(
    here("scratch","bam_readcount","Pacbio_RM_sequel_snv_A.tsv"),
    draft_snvs_af_df, 'snv', 'Pacbio'
)
Pacbio_sequel_snv_B_target <- read_and_annotate_platform(
    here("scratch","bam_readcount","Pacbio_RM_sequel_snv_B.tsv"),
    draft_snvs_af_df, 'snv', 'Pacbio'
)
Pacbio_sequel_indel_A_target <- read_and_annotate_platform(
    here("scratch","bam_readcount","Pacbio_RM_sequel_indel_A.tsv"),
    draft_indels_af_df, 'indel', 'Pacbio'
)
Pacbio_sequel_indel_B_target <- read_and_annotate_platform(
    here("scratch","bam_readcount","Pacbio_RM_sequel_indel_B.tsv"),
    draft_indels_af_df, 'indel', 'Pacbio'
)

# All instruments combined
Pacbio_snv_combined <- summarise_reads(
    Pacbio_revio_snv_target, Pacbio_sequel_snv_A_target, Pacbio_sequel_snv_B_target,
    platform_name = 'Pacbio'
)
Pacbio_indel_combined <- summarise_reads(
    Pacbio_revio_indel_target, Pacbio_sequel_indel_A_target, Pacbio_sequel_indel_B_target,
    platform_name = 'Pacbio'
)
Pacbio_variants_target_combined <- combine_platform_variants(
    Pacbio_snv_combined, Pacbio_indel_combined, 'Pacbio'
)

# Sub-instrument: Revio only
Pacbio_variants_target_combined_revio <- combine_platform_variants(
    rename_platform_prefix(Pacbio_revio_snv_target,   "Pacbio", "Pacbio_revio"),
    rename_platform_prefix(Pacbio_revio_indel_target, "Pacbio", "Pacbio_revio"),
    'Pacbio_revio'
)

# Sub-instrument: Sequel (A + B combined)
Pacbio_snv_sequel_combined <- summarise_reads(
    rename_platform_prefix(Pacbio_sequel_snv_A_target, "Pacbio", "Pacbio_sequel"),
    rename_platform_prefix(Pacbio_sequel_snv_B_target, "Pacbio", "Pacbio_sequel"),
    platform_name = 'Pacbio_sequel'
)
Pacbio_indel_sequel_combined <- summarise_reads(
    rename_platform_prefix(Pacbio_sequel_indel_A_target, "Pacbio", "Pacbio_sequel"),
    rename_platform_prefix(Pacbio_sequel_indel_B_target, "Pacbio", "Pacbio_sequel"),
    platform_name = 'Pacbio_sequel'
)
Pacbio_variants_target_combined_sequel <- combine_platform_variants(
    Pacbio_snv_sequel_combined, Pacbio_indel_sequel_combined, 'Pacbio_sequel'
)
```

## Downloading Illumina Data
```{r}
illumina_indel_extra <- cols(position = 'd', alt_depth = 'd', depth = 'd', mut_freq = 'd')

Illumina_snv_target <- read_and_annotate_platform(
    here("scratch","bam_readcount","Illumina_RM_snv.tsv"),
    draft_snvs_af_df, 'snv', 'Illumina'
)
Illumina_indel_target <- read_and_annotate_platform(
    here("scratch","bam_readcount","Illumina_RM_indel.tsv"),
    draft_indels_af_df, 'indel', 'Illumina',
    extra_col_types = illumina_indel_extra
)
Illumina_variants_target <- combine_platform_variants(
    Illumina_snv_target, Illumina_indel_target, 'Illumina'
)
```

# Ortho Database with check for germline indels among benchmark; as well as characteristics of V3 database
```{r}
drop_join_redundancies <- function(tbl) {
    tbl %>% select(-all_of(annotation_cols), -method, -index)
}

mosaic_variant_database_RM <- draft_variants %>%
    left_join(
        HG002_mut_tbl_combined %>%
            select(CHROM, POS, REF, ALT, Duplex_depth, Duplex_alt_depth,
                   Duplex_mut_freq, Duplex_qual, Duplex_Lower_CI, Duplex_Upper_CI),
        by = c('CHROM','POS','REF','ALT')
    ) %>%
    left_join(drop_join_redundancies(BGI_variants_target),
              by = c('CHROM','POS','REF','ALT')) %>%
    left_join(drop_join_redundancies(Element_variants_target_combined),
              by = c('CHROM','POS','REF','ALT')) %>%
    left_join(drop_join_redundancies(Element_variants_target_combined_long),
              by = c('CHROM','POS','REF','ALT')) %>%
    left_join(drop_join_redundancies(Element_variants_target_combined_standard),
              by = c('CHROM','POS','REF','ALT')) %>%
    left_join(drop_join_redundancies(Pacbio_variants_target_combined),
              by = c('CHROM','POS','REF','ALT')) %>%
    left_join(drop_join_redundancies(Pacbio_variants_target_combined_revio),
              by = c('CHROM','POS','REF','ALT')) %>%
    left_join(drop_join_redundancies(Pacbio_variants_target_combined_sequel),
              by = c('CHROM','POS','REF','ALT')) %>%
    left_join(drop_join_redundancies(Illumina_variants_target),
              by = c('CHROM','POS','REF','ALT')) %>%
    relocate(var_type, var_cat, SomaticEVS, AF, var_id, dup_var_hit, easy_to_map,
             homopolymer_region, germline_insertion, germline_deletion, .after = indel_change) %>%
    mutate(VAF_Category = case_when(
        AF == 0    ~ '0% VAF',
        AF <= 0.01 ~ '0% - 1% VAF',
        AF <= 0.04 ~ '1% - 4% VAF',
        AF < 0.05  ~ '4% - 5% VAF',
        AF < 0.1   ~ '5% - 10% VAF',
        AF < 0.3   ~ '10% - 30% VAF',
        TRUE       ~ '>30% VAF'
    )) %>%
    mutate(SomaticEVS_Category = case_when(
        SomaticEVS == 0 ~ '0_EVS',
        SomaticEVS <= 1 ~ '0to1_EVS',
        SomaticEVS <= 3 ~ '1to3_EVS',
        SomaticEVS < 7  ~ '3to7_EVS',
        TRUE            ~ '7plus_EVS'
    )) %>%
    mutate(Genomic_Context = case_when(
        easy_to_map == 'hit'               & homopolymer_region == 'hit'                      ~ "Easy to Map Region AND Homopolymer Region",
        easy_to_map == 'hit'               & homopolymer_region == 'homopolymer_region_no_hit' ~ "Easy to Map Region BUT Non-Homopolymer Region",
        easy_to_map == 'easy_to_map_no_hit'& homopolymer_region == 'hit'                      ~ "Homopolymer Region BUT Not Easy to Map Region",
        TRUE ~ "Non-Homopolymer Region AND Not Easy to Map Region"
    )) %>%
    mutate(Germline_Indel_Check = case_when(
        germline_insertion == 'hit' & germline_deletion != "hit" ~ "Insertion_Overlap",
        germline_insertion != 'hit' & germline_deletion == "hit" ~ "Deletion_Overlap",
        germline_insertion == 'hit' & germline_deletion == "hit" ~ "ERROR",
        TRUE ~ "No_Overlap"
    )) %>%
    mutate(Coverage_Category_Duplex = case_when(
        Duplex_depth < 100                     ~ '<100X',
        Duplex_depth < 300 & SomaticEVS >= 100 ~ '100X-300X',
        TRUE                                   ~ '>300X'
    )) %>%
    
    mutate(Duplex_Presence   = if_else(Duplex_alt_depth   > 1 & !is.na(Duplex_mut_freq),   'TRUE', 'FALSE')) %>%
    mutate(Element_Presence  = if_else(Element_alt_depth  > 1 & !is.na(Element_mut_freq),  'TRUE', 'FALSE')) %>%
    mutate(BGI_Presence      = if_else(BGI_alt_depth      > 1 & !is.na(BGI_mut_freq),      'TRUE', 'FALSE')) %>%
    mutate(Pacbio_Presence   = if_else(Pacbio_alt_depth   > 1 & !is.na(Pacbio_mut_freq),   'TRUE', 'FALSE')) %>%
    mutate(Illumina_Presence = if_else(Illumina_alt_depth > 1 & !is.na(Illumina_mut_freq), 'TRUE', 'FALSE')) %>%
    
    mutate(Detection_Category = pmap_chr(
        list(Illumina_Presence, Duplex_Presence, Element_Presence, BGI_Presence, Pacbio_Presence),
        function(ill, dup, ele, bgi, pac) {
            methods  <- c(Illumina = ill, Duplex = dup, Element = ele, BGI = bgi, Pacbio = pac)
            detected <- names(methods[methods == 'TRUE'])
            if      (length(detected) == 0) 'NONE'
            else if (length(detected) == 5) 'ALL'
            else    paste(detected, collapse = ',')
        }
    )) %>%
    relocate(VAF_Category, SomaticEVS_Category, Coverage_Category_Duplex,
             Genomic_Context, Detection_Category, .after = dup_var_hit) %>%
    mutate(
        x_ci        = rowSums(select(., Illumina_alt_depth, Duplex_alt_depth, BGI_alt_depth,
                                     Element_alt_depth, Pacbio_alt_depth), na.rm = TRUE),
        n_ci        = rowSums(select(., Illumina_depth, Duplex_depth, BGI_depth,
                                     Element_depth, Pacbio_depth), na.rm = TRUE),
        x_ci_no_Dup = rowSums(select(., Illumina_alt_depth, BGI_alt_depth,
                                     Element_alt_depth, Pacbio_alt_depth), na.rm = TRUE),
        n_ci_no_Dup = rowSums(select(., Illumina_depth, BGI_depth,
                                     Element_depth, Pacbio_depth), na.rm = TRUE)
    )

mosaic_variant_database_RM <- mosaic_variant_database_RM %>%
    cbind(binconf(mosaic_variant_database_RM$x_ci, mosaic_variant_database_RM$n_ci, 0.02)) %>%
    rename(Orthogonal_Lower_CI = Lower, Orthogonal_Upper_CI = Upper) %>%
    select(-PointEst)

mosaic_variant_database_RM <- mosaic_variant_database_RM %>%
    cbind(binconf(mosaic_variant_database_RM$x_ci_no_Dup, mosaic_variant_database_RM$n_ci_no_Dup, 0.02)) %>%
    rename(Ortho_NO_DUP_Lower_CI = Lower, Ortho_NO_DUP_Upper_CI = Upper) %>%
    select(-PointEst)

mosaic_variant_database_RM <- mosaic_variant_database_RM %>%
    mutate(Overlap_Check = case_when(
        Orthogonal_Upper_CI >  Ortho_NO_DUP_Upper_CI & Ortho_NO_DUP_Upper_CI > Orthogonal_Lower_CI  ~ 'Overlap_Detected',
        Orthogonal_Upper_CI <  Ortho_NO_DUP_Upper_CI & Orthogonal_Upper_CI   > Ortho_NO_DUP_Lower_CI ~ 'Overlap_Detected',
        Orthogonal_Upper_CI == Ortho_NO_DUP_Upper_CI & Orthogonal_Lower_CI  == Ortho_NO_DUP_Lower_CI ~ 'Overlap_Detected',
        Orthogonal_Upper_CI == 'NaN' & Ortho_NO_DUP_Upper_CI == 'NaN' ~ 'NA',
        TRUE ~ 'No_Overlap'
    )) %>%
    mutate(Ortho_CI_Flag   = if_else(AF >= Orthogonal_Lower_CI    & AF <= Orthogonal_Upper_CI,    'PASS', 'FAIL')) %>%
    mutate(NonDup_CI_Flag  = if_else(AF >= Ortho_NO_DUP_Lower_CI  & AF <= Ortho_NO_DUP_Upper_CI,  'PASS', 'FAIL'))

write_tsv(mosaic_variant_database_RM, here('scratch','mosaic_variant_database_RM.tsv'))
```

### Sanity Check of Databases
```{r}
mosaic_variant_database_RM %>%
    group_by(Detection_Category) %>%
    summarise(n = n()) %>%
    gt() %>%
    tab_header(title = 'RM Benchmark SNV detection (Cutoffs: Alt_Depth, Mapping Quality, Base Quality)')

mosaic_variant_database_RM %>%
    filter(var_cat == 'candidate') %>%
    group_by(Ortho_CI_Flag) %>%
    dplyr::summarise(count = n()) %>%
    gt() %>%
    tab_header(title = 'RM Benchmark Orthogonal CI Check: Candidate Variants')

mosaic_variant_database_RM %>%
    filter(var_cat == 'putative') %>%
    group_by(Ortho_CI_Flag) %>%
    dplyr::summarise(count = n()) %>%
    gt() %>%
    tab_header(title = 'RM Benchmark Orthogonal CI Check: Putative Variants')

mosaic_variant_database_RM %>%
    filter(var_cat == 'candidate') %>%
    group_by(NonDup_CI_Flag) %>%
    dplyr::summarise(count = n()) %>%
    gt() %>%
    tab_header(title = 'RM Benchmark NonDuplex Orthogonal CI Check: Candidate Variants')

mosaic_variant_database_RM %>%
    filter(var_cat == 'putative') %>%
    group_by(NonDup_CI_Flag) %>%
    dplyr::summarise(count = n()) %>%
    gt() %>%
    tab_header(title = 'RM Benchmark NonDuplex Orthogonal CI Check: Putative Variants')

mosaic_variant_database_RM %>%
    filter(NonDup_CI_Flag == 'FAIL') %>%
    group_by(Detection_Category) %>%
    dplyr::summarise(count = n()) %>%
    gt() %>%
    tab_header(title = 'RM Benchmark NonDuplex Orthogonal CI Check: FAILURES')

mosaic_variant_database_RM %>%
    filter(Ortho_CI_Flag == 'FAIL') %>%
    group_by(Detection_Category) %>%
    dplyr::summarise(count = n()) %>%
    gt() %>%
    tab_header(title = 'RM Benchmark Orthogonal CI Check: FAILURES')

mosaic_variant_database_RM %>%
    group_by(Detection_Category, VAF_Category, Genomic_Context, var_cat, var_type) %>%
    dplyr::summarize(n = n()) %>%
    write_tsv(here('scratch','RM_benchmark_detection_orth_method_genomic_context.tsv'))

mosaic_variant_database_RM %>%
    group_by(Genomic_Context) %>%
    dplyr::summarize(n = n()) %>%
    gt() %>%
    tab_header(title = 'Benchmark SNV detection (alt depth > 1)')
```
